%\VignetteIndexEntry{Boarding-school influenza example}
%\VignetteEngine{knitr::knitr}

# Boarding-School Flu Outbreak Analysis  
***Aaron A. King***

```{r setup,include=FALSE}
require(pomp)
require(plyr)
require(reshape2)
options(stringsAsFactors=FALSE,keep.source=TRUE,encoding="UTF-8")

require(ggplot2)
theme_set(theme_bw())

require(knitr)
opts_knit$set(out.format="html")
opts_chunk$set(
  progress=TRUE,
  prompt=FALSE,tidy=FALSE,highlight=TRUE,
  strip.white=TRUE,
  warning=FALSE,message=FALSE,error=FALSE,
  echo=TRUE,cache=FALSE,
  results='markup',
  fig.show='asis',
  fig.height=5,fig.width=10,
  dpi=100
  )

require(pompExamples)
set.seed(862663052L)
```

First, a little function to cache the results of expensive computations.

```{r juliaChild}
juliaChild <- function (file, expr) {
  if (file.exists(file)) {
    readRDS(file)
    } else {
      val <- eval(expr)
      saveRDS(val,file=file)
      val
      }
  }
```

## Flu model with exponentially-distributed waiting times

```{r bsflu1,cache=FALSE}
pompExample(bsflu)
coef(bsflu)
```
```{r plot1}
simdat <- simulate(bsflu,nsim=10,obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

### Trajectory matching

```{r tm1}
juliaChild("bsflu-tm1.rds",{
  traj.match(bsflu,est=c("Beta","inf.pd","rho"),transform=TRUE) -> tm
  traj.match(tm,method='subplex') -> tm
  data.frame(loglik=logLik(tm),loglik.se=0,as.list(coef(tm)))
}) -> tm1
```
```{r tm1-simplot}
simdat <- simulate(bsflu,params=unlist(tm1),nsim=10,
                   obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

### Iterated filtering

```{r mf1}
juliaChild("bsflu-mf1.rds",{
  mif(bsflu,rw.sd=c(Beta=0.05,inf.pd=0.05,rho=0.05),
      cooling.fraction=0.9,var.factor=2,
      Nmif=50,Np=1000,method='mif2',transform=TRUE) -> mf
  mif(mf,Nmif=50,cooling.fraction=0.5) -> mf
  mif(mf,Nmif=50,cooling.fraction=0.1) -> mf
  ll <- unname(logmeanexp(raply(5,logLik(pfilter(mf))),se=TRUE))
  data.frame(loglik=ll[1],loglik.se=ll[2],as.list(coef(mf)))
}) -> mf1
```
```{r mf1-simplot}
simdat <- simulate(bsflu,params=unlist(mf1),nsim=10,
                   obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

## Model with Erlang(3) confinement period

```{r bsflu3,cache=FALSE}
pompExample(bsflu3)
coef(bsflu3)
```
```{r plot3}
simdat <- simulate(bsflu3,nsim=10,obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

### Trajectory matching

```{r tm3}
juliaChild("bsflu-tm3.rds",{
  traj.match(bsflu3,est=c("Beta","inf.pd","rho"),transform=TRUE) -> tm
  traj.match(tm,method='subplex') -> tm
  data.frame(loglik=logLik(tm),loglik.se=0,as.list(coef(tm)))
}) -> tm3
```
```{r tm3-simplot}
simdat <- simulate(bsflu3,params=unlist(tm3),nsim=10,
                   obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

### Iterated filtering

```{r mf3}
juliaChild("bsflu-mf3.rds",{
  mif(bsflu3,rw.sd=c(Beta=0.05,inf.pd=0.05,rho=0.05),
      cooling.fraction=0.9,var.factor=2,
      Nmif=50,Np=1000,method='mif2',transform=TRUE) -> mf
  mif(mf,Nmif=50,cooling.fraction=0.5) -> mf
  mif(mf,Nmif=50,cooling.fraction=0.1) -> mf
  ll <- unname(logmeanexp(raply(5,logLik(pfilter(mf))),se=TRUE))
  data.frame(loglik=ll[1],loglik.se=ll[2],as.list(coef(mf)))
}) -> mf3
```
```{r mf3-simplot}
simdat <- simulate(bsflu3,params=unlist(mf3),nsim=10,
                   obs=TRUE,as.data.frame=TRUE,include.data=TRUE)
ggplot(data=melt(simdat,id=c("sim","time")),
       mapping=aes(x=time,y=value,group=interaction(variable,sim),
                   color=variable,size=sim=="data",alpha=sim=="data"))+
  geom_line()+
  scale_alpha_manual(values=c(`TRUE`=1,`FALSE`=0.5))+
  scale_size_manual(values=c(`TRUE`=2,`FALSE`=1))
```

## Model comparison

```{r comp}
tb <- ldply(list(det1=tm1,stoch1=mf1,det3=tm3,stoch3=mf3),.id="model")
kable(tb)
```

```{r probes,fig.height=6,fig.width=6}
plist <- list(probe.acf("confined",lags=c(2),type="cov",transform=sqrt),
              probe.ccf(c("confined","convalescent"),lags=c(0,3),transform=sqrt),
              tot=function (x) apply(x,1,sum))
plot(probe(bsflu,params=unlist(tm1),nsim=500,probes=plist))
plot(probe(bsflu,params=unlist(mf1),nsim=500,probes=plist))
plot(probe(bsflu3,params=unlist(tm3),nsim=500,probes=plist))
plot(probe(bsflu3,params=unlist(mf3),nsim=500,probes=plist))
```
